---
title: "Assessment"
output: html_document
date: "2024-11-13"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warnings=FALSE)
```

## How has COVID-19 impacted Amoxicillin prescriptions in Scotland over the last five years, and does this vary across different areas?

First, I will load the needed library packages. I will also load the data and remove any NA's. I will be focusing on data from March as it is close to infection seasons.

```{r, libraries}
library(tidyverse)
library(janitor) # cleaning data
library(gt) # tables
library(here) # directory structure
```

```{r, loading}
march2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/a42762ac-47cb-4fb6-b9b1-2478a588c0ed/download/pitc202403.csv") %>% 
  clean_names()

march2023 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/8dd06c58-1a09-483a-8a01-5d68cfb8b38e/download/pitc202303.csv") %>% 
  clean_names()

march2022 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/a0ec3bf2-7339-413b-9c66-2891cfd7919f/download/pitc202203.csv") %>% 
  clean_names()

march2021 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/df6fc708-5c50-4d57-a5c4-faa19a92c227/download/pitc202103.csv") %>% 
  clean_names()

march2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9581bd8d-5568-4462-93a6-6d7bfbbe7cbc/download/pitc202003.csv") %>% 
  clean_names()

march2019 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/87b21840-beb8-40fd-bad8-579643bc8b8b/download/pitc201903.csv") %>% 
  clean_names()

```

## Visualising the data

I will be aiming to explore how the Amoxicillin prescriptions differ between 2019 and 2024. I will start by looking at the prescribed Amoxicillin in Scotland in from March 2019 (pre-covid) to March 2024

```{r, Amoxicillin, echo=FALSE}

march_datasets <- list(
  march2019 = march2019,
  march2020 = march2020,
  march2021 = march2021,
  march2022 = march2022,
  march2023 = march2023,
  march2024 = march2024
)

amoxicillin_data <- function(data) {
  data %>%
    filter(!is.na(bnf_item_description)) %>%
    filter(str_detect(bnf_item_description, "AMOXICILLIN")) %>%
    group_by(bnf_item_description) %>%
    summarise(quantity_sum = sum(paid_quantity, na.rm = TRUE)) %>%
    arrange(-quantity_sum) %>%
    gt()
}

amoxicillin_tables <- map(march_datasets, amoxicillin_data)

amox19 <- amoxicillin_tables$march2019
amox20 <- amoxicillin_tables$march2020
amox21 <- amoxicillin_tables$march2021
amox22 <- amoxicillin_tables$march2022
amox23 <- amoxicillin_tables$march2023
amox24 <- amoxicillin_tables$march2024
```

## Amoxicillin prescriptions 2019
```{r, Amoxicillin2019, echo=FALSE}
amox19
```

## Amoxicillin prescriptions 2020
```{r, Amoxicillin2020, echo=FALSE}
amox20
```

## Amoxicillin prescriptions 2021
```{r, Amoxicillin2021, echo=FALSE}
amox21
```

2021, very sharp decrease in Amoxicillin prescriptions - could this be linked to the end of lockdown in Scotland and the use of masks?

## Amoxicillin prescriptions 2022

```{r, Amoxicillin202, echo=FALSE}
amox22
```

## Amoxicillin prescriptions 2023
```{r, Amoxicillin2023, echo=FALSE}
amox23
```

## Amoxicillin prescriptions 2024
```{r, Amoxicillin2024, echo=FALSE}
amox24
```

The tables show how Amoxicillin prescriptions increase during lockdown but have a sharp decrease in 2021, when lockdown ended. By looking at the hospital admissions, it could suggest how COVID-19 cases are linked.

I will be looking at hospital admissions from COVID-19
```{r, covid, echo=FALSE}

covid_hospitaladmissions <- read_csv("https://www.opendata.nhs.scot/dataset/578ff074-93f6-4127-a872-0ab97f9d5587/resource/f8f3a435-1925-4c5a-b2e8-e58fdacf04bb/download/hospital_admissions_hb_agesex_20231005.csv") %>% 
  clean_names()

march_dates <- paste0(2020:2024, "03")

admissions_march_data <- purrr::map(march_dates, function(march_date) {
  covid_hospitaladmissions %>% 
  filter(!is.na(hb)) %>% 
  filter(age_group == "All ages", sex == "All", admission_type == "All") %>% 
  filter(str_detect(week_ending, march_date)) %>% 
  select(week_ending, hb, age_group, sex, number_admissions, percent_variation, admission_type) %>%
  group_by(hb) %>% 
  summarise(total_admissions = sum(number_admissions, na.rm = TRUE)) %>% 
  arrange(-total_admissions) %>% 
  gt()
})

names(admissions_march_data) <- paste0("admissions", 2020:2024)

admissions2020 <- admissions_march_data$admissions2020
admissions2021 <- admissions_march_data$admissions2021
admissions2022 <- admissions_march_data$admissions2022
admissions2023 <- admissions_march_data$admissions2023

# there is no data for 2019 or 2024.

```

## Covid2020
```{r, Covid2020, echo=FALSE}
admissions2020
```

## Covid2021
```{r, Covid2021, echo=FALSE}
admissions2021
```

## Covid2022
```{r, Covid2022, echo=FALSE}
admissions2022
```
## Covid2023
```{r, Covid2023, echo=FALSE}
admissions2023
```

In the report I will aim to explore if mask use due to COVID-19 decreased the need for antibiotic prescriptions. And if the increase in antibiotics prescriptions (higher than before COVID-19) could be due to a weakened immune system from COVID-19 leading to a higher susceptibility of having bacterial infections. I will also be exploring if there is a link between the number of cases (e.g. Hospital admissions data from COVID-19) is related to higher or lower antibiotic use in that area post-COVID-19.

My next step will be creating a Bar plot showing both COVID-19 admissions and Amoxicillin prescriptions. I will be focusing as well on the different healthboards - if there is a link between COVID-19 awareness and restrictions with Amoxicillin prescriptions (taking into account the population number). 